---
title: "The Midwest from Ohio"
author: "Kevin Reuning"
format: 
    clean-revealjs:
        width: 1600
        height: 900
        slide-number: false
        callout-icon: false
        execute:
            echo: false
            fig-align: "center"
            warning: false
---


## Research Questions

- How is the midwest regionally defined? 
- Who identifies as midwesterner? 
- Where is the midwest going? 

## Survey

## Data 

```{r}
library(here)
library(tidyverse)
library(srvyr)
library(usmap)
library(rlang)
library(purrr)
library(gt)
library(scales)
library(plotly)
library(usmap)
library(ggrepel)
library(huxtable)

get_data <- function(variable, data) {
  data |>
    group_by(!!rlang::sym(variable)) |>
    summarize(survey_mean()) |>
    rename("Response" = !!rlang::sym(variable)) |>
    mutate(qname = variable) |>
    mutate(
      Response = factor(
        Response,
        levels = c(
          "On",
          "Off",
          "Very representative",
          "Somewhat representative",
          "Not very representative",
          "Not at all representative",
          "Very accepting",
          "Accepting",
          "Somewhat accepting",
          "Not very accepting",
          "Not accepting at all",
          "Strongly agree",
          "Agree",
          "Somewhat agree",
          "Neither agree nor disagree",
          "Somewhat disagree",
          "Disagree",
          "Strongly disagree",
          "A great deal of difference",
          "A big difference",
          "A small difference",
          "No difference",
          "0-10 minutes",
          "11-20 minutes",
          "21-30 minutes",
          "31-45 minutes",
          "Yes",
          "No",
          "Less than 1",
          "1-5 years",
          "6-15 years",
          "16-30 years",
          "30-50 years",
          "50+ years",
          "City",
          "Suburb",
          "Town",
          "Rural Area",
          "Democrat",
          "Republican",
          "Independent",
          "Donald Trump (Republican)",
          "Kamala Harris (Democrat)",
          "Someone else",
          "Couldn't vote",
          "Didn't vote",
          "Don't remember",
          "Man",
          "Woman",
          "Non-binary / third gender",
          "Less than $25,000",
          "$25,000-$49,999",
          "$50,000-$74,999",
          "$75,000-$99,999",
          "$100,000-$149,999",
          "$150,000 or more",
          "Some high school or less",
          "High school diploma or GED",
          "Some college, but no degree",
          "Associates or technical degree",
          "Bachelor’s degree",
          "Graduate or professional degree (MA, MS, MBA, PhD, JD, MD, DDS etc.)",
          "Prefer not to say",
          "Other",
          "Not Sure",
          "Not sure",
          "Skipped"
        )
      )
    ) |>
    arrange(Response)
}


data <- read_csv(here("surveys", "2025", "survey_data_cleaned.csv"))

svy_data <- as_survey(data, weights = weights)
vars <- read_csv(here("surveys", "2025", "variables.csv"))

n <- nrow(data)


```


## Identity and Partisanship 



```{r}

all <- get_data("Q10", svy_data) |>
  mutate(Type = "All")
dem <- get_data("Q10", filter(svy_data, grepl("Democrat", PID7))) |>
  mutate(Type = "Dem")
rep <- get_data("Q10", filter(svy_data, grepl("Republican", PID7))) |>
  mutate(Type = "Rep")

rbind(all, dem, rep) |>
  pivot_wider(id_cols = Type, values_from = coef, names_from = Response) |>
  hux() |>
  set_number_format(row = 2:4, value = fmt_percent(digits = 0)) |>
  set_bottom_border(row = 1, col = everything())

```


# What is the Midwest? 

## Midwest from Ohio? 



```{r}

plot_data <- filter(data) |>
  select(starts_with("Q3_")) |>
  mutate(across(Q3_1:Q3_29, ~ .x == "On")) |>
  summarize(across(Q3_1:Q3_29, ~ weighted.mean(.x, w = data$weights))) |>
  pivot_longer(everything()) |>
  left_join(vars, by = join_by(name == qname)) |>
  select(sub, value) |>
  mutate(fips = fips(sub))


map_data <- us_map(region = "states", include = plot_data$sub) |>
  left_join(plot_data, by = join_by(full == sub))

ggplot(data = (map_data), aes(fill = value)) +
  geom_sf() +
  scale_fill_viridis_c(
    "Midwest?",
    option = "B",
    end = 0.8,
    labels = scales::label_percent()
  ) +
  geom_sf_text(
    aes(label = scales::percent(value, accuracy = 1)),
    color = "white",
    size = 6
  ) +
  theme_void() +
  theme(legend.position = "bottom")

```


## Partisan Differences

```{r}

dem_data <- filter(data, grepl("Democrat", PID7))
dem_map <- dem_data |>
  select(starts_with("Q3_")) |>
  mutate(across(Q3_1:Q3_29, ~ .x == "On")) |>
  summarize(across(Q3_1:Q3_29, ~ weighted.mean(.x, w = dem_data$weights))) |>
  pivot_longer(everything()) |>
  left_join(vars, by = join_by(name == qname)) |>
  select(sub, value) |>
  mutate(fips = fips(sub)) |>
  rename(Dem = value)


rep_data <- filter(data, grepl("Republican", PID7))
rep_map <- rep_data |>
  select(starts_with("Q3_")) |>
  mutate(across(Q3_1:Q3_29, ~ .x == "On")) |>
  summarize(across(Q3_1:Q3_29, ~ weighted.mean(.x, w = rep_data$weights))) |>
  pivot_longer(everything()) |>
  left_join(vars, by = join_by(name == qname)) |>
  select(sub, value) |>
  mutate(fips = fips(sub)) |>
  rename(Rep = value)


partisan_map_data <- left_join(rep_map, dem_map)

```



```{r}
#| layout-ncol: 2

partisan_map_data |>
  ggplot(aes(x = Rep, y = Dem)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_point(size = 4, alpha = 0.5) +
  geom_text_repel(aes(label = sub), size = 4) +
  theme_minimal() +
  scale_x_continuous(labels = scales::label_percent()) +
  scale_y_continuous(labels = scales::label_percent())


partisan_map_data |>
  mutate(Diff = Dem - Rep) |>
  left_join(us_map(region = "states"), by = join_by(sub == full)) |>
  ggplot(aes(fill = Diff, geometry = geom)) +
  geom_sf() +
  scale_fill_gradient2(
    "More Dem?",
    midpoint = 0,
    low = "orangered4",
    mid = "grey80",
    high = "steelblue4",
    labels = scales::label_percent()
  ) +
  geom_sf_text(
    aes(label = scales::percent(Diff, accuracy = 1)),
    color = "white",
    size = 6
  ) +
  theme_void() +
  theme(legend.position = "bottom")

```

# Who is Midwestern? 

## Self-Identity 

```{r}

no_data <- filter(data, grepl("No", Q10))
no_data <- no_data |>
  select(starts_with("Q3_")) |>
  mutate(across(Q3_1:Q3_29, ~ .x == "On")) |>
  summarize(across(Q3_1:Q3_29, ~ weighted.mean(.x, w = no_data$weights))) |>
  pivot_longer(everything()) |>
  left_join(vars, by = join_by(name == qname)) |>
  select(sub, value) |>
  mutate(fips = fips(sub)) |>
  rename(Not = value)


yes_data <- filter(data, grepl("Yes", Q10))
yes_data <- yes_data |>
  select(starts_with("Q3_")) |>
  mutate(across(Q3_1:Q3_29, ~ .x == "On")) |>
  summarize(across(Q3_1:Q3_29, ~ weighted.mean(.x, w = yes_data$weights))) |>
  pivot_longer(everything()) |>
  left_join(vars, by = join_by(name == qname)) |>
  select(sub, value) |>
  mutate(fips = fips(sub)) |>
  rename(Midwesterner = value)


identity_data <- left_join(yes_data, no_data)


```


```{r}
#| layout-ncol: 2

identity_data |>
  ggplot(aes(x = Midwesterner, y = Not)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_point(size = 4, alpha = 0.5) +
  geom_text_repel(aes(label = sub), size = 4) +
  theme_minimal() +
  scale_x_continuous(labels = scales::label_percent()) +
  scale_y_continuous(labels = scales::label_percent())


identity_data |>
  mutate(Diff = Midwesterner - Not) |>
  left_join(us_map(region = "states"), by = join_by(sub == full)) |>
  ggplot(aes(fill = Diff, geometry = geom)) +
  geom_sf() +
  scale_fill_gradient2(
    "Midwesterner\nMore Likely?",
    midpoint = 0,
    low = "orangered4",
    mid = "grey80",
    high = "steelblue4",
    labels = scales::label_percent()
  ) +
  geom_sf_text(
    aes(label = scales::percent(Diff, accuracy = 1)),
    color = "white",
    size = 6
  ) +
  theme_void() +
  theme(legend.position = "bottom")

```

## Values of the Midwest


```{r}

value_vars <- c("Q4_1", "Q4_2", "Q4_3", "Q4_4", "Q4_5")
out <- map(value_vars, get_data, data = svy_data)
names(out) <- value_vars
out <- list_rbind(out, names_to = "qname") |> left_join(vars)


out |>
  pivot_wider(id_cols = sub, names_from = Response, values_from = coef) |>
  hux() |>
  set_number_format(row = 2:6, value = fmt_percent(digits = 0)) |>
  set_contents(row = 1, col = 1, value = "")

```



## Partisan Differences


```{r}

value_vars <- c("Q4_1", "Q4_2", "Q4_3", "Q4_4", "Q4_5")
out_rep <- map(
  value_vars,
  get_data,
  data = filter(svy_data, grepl("Republican", PID7))
)
names(out_rep) <- value_vars
out_rep <- list_rbind(out_rep, names_to = "qname") |> left_join(vars)

out_dem <- map(
  value_vars,
  get_data,
  data = filter(svy_data, grepl("Democrat", PID7))
)
names(out_dem) <- value_vars
out_dem <- list_rbind(out_dem, names_to = "qname") |> left_join(vars)

out_dem$Party <- "Democratic"
out_rep$Party <- "Republican"

rbind(out_dem, out_rep) |>
  pivot_wider(
    id_cols = c(Response, sub),
    names_from = Party,
    values_from = coef
  ) |>
  mutate(Diff = Democratic - Republican) |>
  pivot_wider(id_cols = sub, names_from = Response, values_from = Diff) |>
  hux() |>
  set_number_format(row = 2:6, value = fmt_percent(digits = 0)) |>
  map_text_color(
    -(1:2),
    -1,
    by_colorspace("orangered4", "grey20", "steelblue4")
  ) |>
  set_contents(row = 1, col = 1, value = "")

```

# Direction of the Midwest?

## Direction of Midwest 

Agree/Disagree:  "The Midwest is better off today than it was 25 years ago."

```{r}

value_vars <- c("Q7")
out_dir <- map(value_vars, get_data, data = svy_data)
names(out_dir) <- value_vars
out_dir <- list_rbind(out_dir, names_to = "qname") |> left_join(vars)

out_dir_dem <- map(
  value_vars,
  get_data,
  data = filter(svy_data, grepl("Democrat", PID7))
)
names(out_dir_dem) <- value_vars
out_dir_dem <- list_rbind(out_dir_dem, names_to = "qname") |> left_join(vars)

out_dir_rep <- map(
  value_vars,
  get_data,
  data = filter(svy_data, grepl("Republican", PID7))
)
names(out_dir_rep) <- value_vars
out_dir_rep <- list_rbind(out_dir_rep, names_to = "qname") |> left_join(vars)


out_dir$Type <- "All"
out_dir_dem$Type <- "Dem"
out_dir_rep$Type <- "Rep"

rbind(out_dir, out_dir_dem, out_dir_rep) |>
  pivot_wider(id_cols = Type, values_from = coef, names_from = Response) |>
  hux() |>
  set_number_format(row = 2:4, value = fmt_percent(digits = 0)) |>
  set_contents(row = 1, col = 1, value = "Group")

```

